Refactoring of Software Using Relational Database

### 2013-02-13 doc/paper.tex
new style, formatting, changes in related work

\documentclass[11pt]{article}
\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{semantic}
\usepackage{caption}
\usepackage{subcaption}

\title{Refactoring of Software Using Relational Database}
\author{Ondrej Macek}

\begin{document}
\maketitle
\section{Introduction}
\label{sec:intro}
The evolution (change) of a software is a common issue during software development. It occurs from many reasons in all phases of software lifecycle. The evolution severity usually depends on number of changes which has to be made and on the number of affected software components. The focus of this paper is in data evolution context of software implemented in object oriented language using relational database. It means that the data evolution affect at least two software components: persistent objects (entities) and the database. The evolution is based on developer's point of view in this paper as it origins in the need to evolve code. This kind of evolution is straightforward in case there are no data stored in the database, because the database schema can be generated according to an evolved code, moreover some object-relational mapping (ORM) frameworks are able to process basic evolutionary transformation on database with stored data. The evolution of an software with ORM become more difficult when there are stored data and when the evolutionary scenario is more complex. The data migration to the new database schema is usually processed manually, which is expensive and error prone.

@@

\section{Problem Definition} 
\label{sec:problem}
% víc rozepsat o změnách a jak se jim bránit a kam tím směřujeme
\textbf{TODO: write an introduction when and why the software changes. and possible solutions.}
A software implemented using an object-oriented language, which uses relational database as a data storage (as illustrated in the Figure \ref{fig:evolution}) is object of interest of this paper. There are three important components of such a software: application (its persistent layer concretely), database (consisting of database schema and stored data) and an ORM, therefore we define a software as a triple consisting of an application and a database which are connected together by an ORM:
\begin{align*}
& \mathbf{software} = ( Application, Database, ORM )
\end{align*}
The software has to be in consistent state so its users could have benefit from its usage. The state of the software is defined to communicate the sate of the software to its user, the symbol $\perp$ is used as a notation for inconsistent state. 
\begin{align*}
& \mathbf{state} = Consistent\; s \: | \: \perp, s \in State 
\end{align*}
The software is in consistent state if application and database are consistent and the database structure corresponds to the application structure according to the ORM.

The evolution of a software is a transformation from one consistent state to another one. These states are called generations of the software and the function which changed the state of a software are called transformations. The evolution consists of change in application and in database as shown in the Figure \ref{fig:evolution}. 
\begin{figure}
\centering
	\includegraphics[scale=0.4]{./images/evolution_simple}
	\caption{The evolution of data changes the system on all levels. The figure shows all the components of the evolution process.}
	\label{fig:evolution}
\end{figure}
The usual approach is to refactor code (which is quite fast due to help of IDE) and then re-generation or manual migration of the database. Another approach is to evolve a database (e.g. due to database optimizations) and then update the code. This approach is very common too, nevertheless this paper focus on evolution which origin from the application point of view.

The refactoring itself does not provide enough information for the complex migration of database and data, therefore a level defining the evolution is added - the set of all possible transformations $E$. The evolution definition from this level is propagated into both involved levels - application and database. The evolution is defined as:
\begin{align*}
& t(state(s)) = \begin{cases}
state(sw(\Psi(s, t), \Phi(s,t)) \; \mathbf{if} \; state(s) = Consistent \; s\\\\
 	\perp
\end{cases}\\ 
& s \in Software, t \in E
\end{align*}
Where $\Psi$ maps the software evolution cases to the code refactoring and $\Phi$ to maps to the evolutionary transformation of database.

@@

\paragraph{Tools for Database Evolution} The MeDEA project \cite{Meda:Main} offers a tool for evolution of both database schema and stored data based on model-driven approach. The PRISM is a research project for data management under schema evolution \cite{PRISM} in contrast with our proposal, and with MeDEA, it extends the SQL command set by so-called schema modification operators which implements the schema evolution.

The project DB-MAIN \cite{Hick:DBMAIN} provides a MDD approach to data evolution on all the levels of software we do. The project is well documented formally. The difference between DB-MAIN and our framework is that each has a different focus. DB-MAIN starts with database modification and database migration, and entities are created accordingly; whereas our focus is on entities evolution, which is then propagated to a database with an emphasis on automatization. Our goal is to hide the entire database level from our users.

All mentioned project is aimed to  be used by database administrators

\paragraph{ORM Frameworks} here are many object-relational mapping frameworks available for developers, and some of them provide tools for database migration. Hibernate \cite{Hibernate} is one of the most popular ORM frameworks in the Java community.  It provides customizable ORM for a wide range of databases, however it does not support complex database migrations.  It is capable only of creating a new table or adding a new column, hence it is not possible to, for example, drop a table or copy values from one column to another. Active Record \cite{Active_Record} is an ORM framework in the Ruby on Rails environment. Since its first version it has contained support for database evolution according to the create-update-delete principle, in the form of so-called migrations \cite{Rails:Migrations} which can be extended by adding user SQL commands. Entity Framework \cite{Entity_Framework} is Microsoft's ORM solution for the .NET platform. Its capabilities of data evolution support are similar to those of Active Record.

Very promising solution is Liquibase \cite{Liquibase} a tool for database refactoring and evolution. It is capable to migrate both database schema and stored data. The evolution is described in a form of a XML document which can be interpreted on various databases. The difference is we try not to bother the developer with database-specific details and make all changes according to a change in application code.

@@

\subsection{Application Model}
%TODO popis
\paragraph{Application} Application is defined as a sequence of classes and it creates the context for all structures used in the software persistent layer.
\begin{align*}
& \mathbf{Application} = (Class*) \\
& 	\mathbf{Class} = (label, Property*, Association*) \\
& \mathbf{Property} = (label, AppType, DefaultValue, Cardinality, Mandatory) \\
&	\mathbf{Association} = (label, classRef, startCardinality, endCardinality)  \\
& \mathbf{AppType} = APPSTRING
\end{align*}

@@

\subsubsection{Application Manipulation}
\label{sec:app-evolution}
The application model defines only the structure of an application thus there are defined transformation for its creating, altering and disassembling. Therefore the conditions for successful transformations are simple 1) when creating an object in the model or altering its name collisions has to be prevent, 2) when renaming a class the references have to be updated 3) when creating or altering an association the existence of referenced class has to be verified. The only issue during disassembling is the class can be removed only in case it is not associated by other classes. If a transformation cannot be processed the state of the application become inconsistent. The transformations are in table \ref{tab:app-evolution}. The list of altering operations is not complete as there are missing transformation for changing the obligation or cardinality of properties and associations. It is because these transformations are not so important for the next text when the complex refactorings are considered. 
\begin{table}
\centering
	\begin{tabular}{|l|}
	\hline
	\textbf{Application Creation} \\
	newProperty(className, property, application) \\
	newAssociation(className, association, mapping, application) \\
	newClass(class, application) \\
	\textbf{Application Alternation} \\
	renameProperty(className, oldName, newName, application) \\
	renameAssociation(className, oldName, newName, application) \\
	renameClass(oldName, newName, application)\\
	changeReference(referncedClass, newReferencedClass) \\
	\textbf{Application Disassembly} \\
	removeProperty(className, propertyName, application) \\
	removeAssociation(className, associationName, application) \\
	removeClass(className, application)\\
	\textbf{Copy Structure}\\
	copyProperty(sourceClass, property, targetClass, newName, application)  \\
	copyClass(class, newName, application) \\
	\hline
	\end{tabular}
	\caption{The transformations for application evolution.}
	\label{tab:app-evolution}
\end{table}

@@

\begin{align*}
&	\mathbf{Database} = ( Table*, Row* )\\
&	\mathbf{Table} = (label, primaryKey, Column*, ForeignKey*)\\
&	\mathbf{Column} = (label, DbType, DefaultValue, Constraint*) \\
&	\mathbf{ForeignKey} = (Label, Label, Constraint*) \\
&	\mathbf{PrimaryKey} =  ( Label, Sequence ) \\
&	\mathbf{DbType} = DBSTRING \; | \; DBINT\\
&	\mathbf{Constraint} = NOTNULL \; | \; UNIQUE 
\end{align*}

@@

A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of value pairs, which represents concrete value of concrete column or key. Each row contains a name of a table it belongs to.
\begin{align*}
&	\mathbf{Row} = (Label, N, Pair*) \\
&	\mathbf{Pair} = (Label, Value) 
\end{align*}

@@

\subsubsection{Database Manipulation}
\label{sec:db-evolution}
The database consists of two parts - from database schema which defines the structure and from stored data - hence the transformations have to consider both parts. The transformation for manipualtion of structure has similar conditions to the evolution of application application. The transformations for data manipulation are inspired by the SQL language. The basic transformations are in Table \ref{tab:db-basic-evolution}. 
\begin{table}
\centering
	\begin{tabular}{|l|}
	\hline
	\textbf{Database Creation} \\
	addColumn(tableName, column, database) \\
	addForeignKey(tableName, key, database) \\
	addTable(table, database)\\
	\textbf{Database Alternation} \\
	alterColumnName(tableName, oldName, newName, dataabse) \\
	alterForeignKeyName(tableName, oldName, newName, dataabse) \\
	alterTableName(oldName, newName, dataabse) \\
	\textbf{Database Disasembling} \\
	dropColumn(tableName, columnName, database) \\
	dropForeignKey(tableName, keyName, database) \\
	dropTable(tableName, database) \\
	\textbf{Data Manipulation} \\
	select(tableName, rowId, database) \\
	select(tableName, database) \\
	insert(row, database) \\
	insert(rowId, value, database) \\
	delete(tableName, rowId, database) \\
	delete(tableName, rowId, columnName, database) \\
	\textbf{Copy Structure} \\
	copyColumnStructure(sourceTableName, targetTableName, \\  \hspace{0.5in} sourceColumnName, newName, database) \\
	copyTableStructure(sourceTable, newTableName, database) \\
	\textbf{Copy Strucutre and Values} \\
	copyColumn(sourceTable, targetTable, \\
	\hspace{0.5in} sourceColumnName, targetColumnName, \\ 
	\hspace{0.5in} mapping, database) \\
	copyTable(sourceTable, newName, database) \\
	\hline
	\end{tabular}
	\caption{The transformations for database evolution.}
	\label{tab:db-basic-evolution}
\end{table}
The set of transformations for database manipulation is limited in contrast with the SQL language it is because mentioned transformation are necessary for application of data evolution on database level.

@@

The mapping between instances of the classes (stored data) to provides information which value from the source table rows will be copied to which row in target table. The value of a column is then inserted for each row into the copied column during $copyColumn$ transformation. 
The mapping is usually defined as: 
\begin{align*}
&	mapping =  (r_1, r_2)*,\\
& 	r_1 \in select(ORM(sourceName), database) \cup \varnothing, \\
& 	r_2 \in select(ORM(targetName), database) \cup \varnothing
\end{align*}
The mapping is not limited to the bijection and it is possible to not include a row into a mapping - this has to be verified because of possible $NOTNULL$ constraints on a target column, as well as the $UNIQUE$ constraint has to be verified because the mapping can violate it in general. On the other hand the benevolent mapping help with creating various kinds on advanced evolutionary transformations. The mapping can cause data losses in some cases if used as a part of advanced refactoring. This situation is discussed later on.

\paragraph{Data Secure Disasembling of a Software} The transformation for disassembly the database can have fatal impact on the data preservation, therefore the set of transformations is extended by a data secure transformations for database disassembling. These transformations are not part of the SQL standard, although they can be implemented as database functions. These transformations creates a secure way to disassembly the database as they drops empty structural elements only. These transformations are in table \ref{tab:db-adv-evolution}. 
\begin{table}
\centering
	\begin{tabular}{|l|}
	\hline
	\textbf{Data Secure Database Disassembling} \\
	dropEmptyColumn(tableName, columnName, database) \\
	dropEmptyForeignKey(tableName, keyName, database) \\
	dropEmptyTable(tableName, database) \\
	\hline
	\end{tabular}
	\caption{The data secure transformations for database evolution which can be created as SQL functions.}
	\label{tab:db-adv-evolution}
\end{table}

@@

\paragraph{Basic Software Evolution} The basic evolutionary transformations of the software are based on basic evolution of an application. These transformation cannot be  mapped directly on the evolutionary transformations of the database, because properties and associations can be mapped as a column (foreign key respectively) or as a table, this has to be reflected during the evolution as shown on the example of creating a new property: 
\begin{align*}
& \Phi(newProperty(className, property, application)) = \\
& \;\;\; = \begin{cases}
  addColumn(ORM(className), ORM(property), database) \\\mathbf{if} \; cardinality(property \leq 1)  \\\\
  addTable(ORM(property), database) \\
  \mathbf{if} \; cardinality(property > 1)  
 \end{cases}
\end{align*}

@@

\begin{table}
\centering
	\begin{tabular}{|p{0.4\textwidth} p{0.5\textwidth}|}
	\hline
	Application Transformation & Database Transformation \\
	\multicolumn{2}{|c|}{\textbf{Software Creation}} \\
	newProperty & addColumn  ($cardinality \leq 1$)\\
	& addTable \\ 
	& \\
	newAssociation & addColumn ($cardinality \leq 1$)\\
	& addTable \\
	& \\
	newClass & addTable \\
	\multicolumn{2}{|c|}{\textbf{Application Alternation}} \\
	renameProperty & renameColumn ($cardinality \leq 1$)\\
	& renameTable \\
	& \\
	renameAssociation & renameForeignKey ($cardinality \leq 1$)\\
	& renameTable \\
	& \\
	renameClass & renameTable \\
	\multicolumn{2}{|c|}{\textbf{Application Disassembly}} \\
	removeProperty & removeColumn ($cardinality \leq 1$)\\
	& removeTable \\
	& \\
	removeAssociation & removeForeignKey ($cardinality \leq 1$)\\
	& removeTable \\
	& \\
	removeClass & removeTable\\
	\hline
	\end{tabular}
	\caption{The mapping between evolution of the application to the database transformations.}
	\label{tab:sw-basic-evolution}
\end{table}

@@

The  $copyProperty$ is the first transformation when an additional information has to be added to the usual code refactoring. It is because the $copyColumn$ transformation needs one more information to succeed - there has to be provided mapping between the source and target table to assure the data information consistency. %Moreover the source and target should have the same structure, type and constraints.
\begin{align*}
& \mathbf{copyProperty}(sourceClass, propertyName, targetClass, \\
& \; \; \; mapping, software) 
\end{align*}
The transformation is interpreted for both software components - in case of the application a new property is added:
\begin{align*}
& \Psi(copyProperty(sourceClass, property, targetClass, mapping, \\ 
&  \; \; \; software)) = newProperty(targetClass, property, application) 
\end{align*}
in case of database the copy of a column or table is created according to property's cardinality and then values are copied:
\begin{align*}
& \Phi(copyProperty(sourceClass, property, targetClass, newName,  \\ 
&  \; \; \; mapping, software)) = \begin{cases}
 coypColumn(ORM(sourceClass),  \\ \;\;\; ORM(targetClass),  \\ \;\;\;alterColumnName(ORM(property), \\ \;\;\; ORM(newName)), mapping, database) \\ \; \mathbf{if} \; cardinality(property) \leq 1 \\\\
 copyTable(ORM(Property), \\ \;\;\; ORM(newName))
 \end{cases}
\end{align*}

@@

\paragraph{Move Property}
The $moveProperty$ transformation is based on the $copy$ $Property$ transformation followed by the $removeProperty$ so its construction is easy. On the other hand special attention has to be paid to the provided mapping of instances, because it can cause data loses. The ideal case is when the mapping is a bijection among instances, then the transformation can not cause the lost of data. On the other hand the situation when the mapping does not provides pairs for all rows in a source table can lead to data loss. 
\begin{align*}
& moveProperty(sourceClass, property, targetClass, newName, \\
& \;\;\; mapping, software) = removeProperty(sourceClass, property, \\
& \;\;\; CopyProperty(sourceClass, property, targetClass, \\ 
& \;\;\;newName, mapping, software))
\end{align*}

\paragraph{Merge Classes}
There occur situations when there is a need to merge two classes and the inlining transformation cannot be used because of name collisions among properties in such a case the $merge$ transformation can be used. It can be used in case structure of both classes (respectively tables) are equal. The transformation creates new rows in the new table - the primary keys from origin instances are replaced by a new ones.
\begin{align*}
& merge(firstClass, secondClass, newClassName, software) = \\
& \; \; \; = removeClass(secondClass, removeClass(firstClass, \\
& \; \; \; changeReference(secondClass, newClassName, \\
& \; \; \; changeReference(firstClass, newClassName, \\
& \; \; \; copyClass(firstClass, newClassName, software))))))
\end{align*}

\paragraph{Extract and Inline Class}
The extract and inline class are two opposite transformations. First of them extract a class from an existing class, second of them moves data from one class into another.
\begin{align*}
& extract(sourceClass, newClass, property*, software) = \\ 
& \; \; \; = moveProperty(sourceClass, newClass, property*, \\ 
& \; \; \; addClass(newClass, software))
\\\\
& inline(targetClass, classToInline, software)) = \\
& \; \; \; = removeClass(classToInline, moveProperty(classToInline, \\
& \; \; \; targetClass, properties(classToInline), software))
\end{align*}

@@

@@

\bibliographystyle{plain}      % basic style, author-year citations
\bibliography{references}   % name your BibTeX data base

>>>>

\documentclass[runningheads]{comsis}

\def\journalissue{ComSIS Vol.~0, No.~0, December 2009}
\def\paperidnum{UDC, DOI: N/A}
\setcounter{page}{1}

\usepackage[dvipdfm]{hyperref}

\usepackage{amssymb}
\usepackage[fleqn]{amsmath}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{semantic}
\usepackage{caption}
\usepackage{subcaption}

\title{Refactoring of Software Using Relational Database}
\author{Ondrej Macek\inst{1}}
\authorrunning{Ondrej Macek}
\institute{Czech Technical University in Prague\\
  Karlovo namesti 13 \\
121 35 Praha 2, Czech Republic\\
  \email{ondrej.macek@fel.cvut.cz}}

\begin{document}
\maketitle

\begin{abstract}

\vspace{6pt}\textbf{Keywords:}
\end{abstract}

\section{Introduction}
\label{sec:intro}
The evolution (change) of a software is a common issue during software development. It occurs from many reasons in all phases of software lifecycle. The evolution severity usually depends on number of changes which has to be made and on the number of affected software components. The focus of this paper is in data evolution context of software implemented in object oriented language using relational database. It means that the data evolution affect at least two software components: persistent objects (entities) and the database. The evolution is based on developer's point of view in this paper as it origins in the need to evolve code. This kind of evolution is straightforward in case there are no data stored in the database, because the database schema can be generated according to an evolved code, moreover some object-relational mapping (ORM) frameworks are able to process basic evolutionary transformation on database with stored data. The evolution of an software with ORM become more difficult when there are stored data and when the evolutionary scenario is more complex. The data migration to the new database schema is usually processed manually, which is expensive and error prone.

@@

\section{Problem Definition} 
\label{sec:problem}
% víc rozepsat o změnách a jak se jim bránit a kam tím směřujeme
\textbf{TODO: write an introduction when and why the software changes. and possible solutions.}
A software implemented using an object-oriented language, which uses relational database as a data storage (as illustrated in the Figure \ref{fig:evolution}) is object of interest of this paper. There are three important components of such a software: application (its persistent layer concretely), database (consisting of database schema and stored data) and an ORM, therefore we define a software as a triple consisting of an application and a database which are connected together by an ORM:
\begin{align}
& \mathbf{software} = ( Application, Database, ORM )
\end{align}
The software has to be in consistent state so its users could have benefit from its usage. The state of the software is defined to communicate the sate of the software to its user, the symbol $\perp$ is used as a notation for inconsistent state. 
\begin{align}
& \mathbf{state} = Consistent\; s \: | \: \perp, s \in State 
\end{align}
The software is in consistent state if application and database are consistent and the database structure corresponds to the application structure according to the ORM.

The evolution of a software is a transformation from one consistent state to another one. These states are called generations of the software and the function which changed the state of a software are called transformations. The evolution consists of change in application and in database as shown in the Figure \ref{fig:evolution}. 
\begin{figure}
\centering
	\includegraphics[width=.7\textwidth]{./images/evolution_simple}
	\caption{The evolution of data changes the system on all levels. The figure shows all the components of the evolution process.}
	\label{fig:evolution}
\end{figure}
The usual approach is to refactor code (which is quite fast due to help of IDE) and then re-generation or manual migration of the database. Another approach is to evolve a database (e.g. due to database optimizations) and then update the code. This approach is very common too, nevertheless this paper focus on evolution which origin from the application point of view.

The refactoring itself does not provide enough information for the complex migration of database and data, therefore a level defining the evolution is added - the set of all possible transformations $E$. The evolution definition from this level is propagated into both involved levels - application and database. The evolution is defined as:
\begin{align}
& t(state(s)) = \begin{cases}
state(sw(\Psi(s, t), \Phi(s,t)) \; \mathbf{if} \; state(s) = Consistent \; s\\\\
 	\perp
\end{cases}\\ 
& s \in Software, t \in E \nonumber
\end{align}
Where $\Psi$ maps the software evolution cases to the code refactoring and $\Phi$ to maps to the evolutionary transformation of database.

@@

@@

\subsection{Application Model}
%TODO popis
\paragraph{Application} Application is defined as a sequence of classes and it creates the context for all structures used in the software persistent layer.
\begin{align}
& \mathbf{Application} = (Class*) \\
& 	\mathbf{Class} = (label, Property*, Association*) \\
& \mathbf{Property} = (label, AppType, DefaultValue, \nonumber \\ & \;\;\;\;  Cardinality, Mandatory) \\
&	\mathbf{Association} = (label, classRef, startCardinality, endCardinality)  \\
& \mathbf{AppType} = APPSTRING
\end{align}

@@

\subsubsection{Application Manipulation}
\label{sec:app-evolution}
The application model defines only the structure of an application thus there are defined transformation for its creating, altering and disassembling. Therefore the conditions for successful transformations are simple 1) when creating an object in the model or altering its name collisions has to be prevent, 2) when renaming a class the references have to be updated 3) when creating or altering an association the existence of referenced class has to be verified. The only issue during disassembling is the class can be removed only in case it is not associated by other classes. If a transformation cannot be processed the state of the application become inconsistent. The transformations are in table \ref{tab:app-evolution}. The list of altering operations is not complete as there are missing transformation for changing the obligation or cardinality of properties and associations. It is because these transformations are not so important for the next text when the complex refactorings are considered. 
\begin{table}
\caption{The transformations for application evolution.}
	\label{tab:app-evolution}
\centering
	\begin{tabular}{ll}
	\hline
	Type of transformation & Defined transformations \\
	\hline
	\\[-2ex] 
	\textbf{Application Creation} 
	& newProperty(className, property, application) \\
	& newAssociation(className, association, mapping, \\ &  \hspace{0.5in}application) \\
	&  newClass(class, application) \\
	\textbf{Application Alternation} 
		& renameProperty(className, oldName, newName, \\ &  \hspace{0.5in} application) \\
		& renameAssociation(className, oldName, newName, \\ &  \hspace{0.5in}application) \\
		& renameClass(oldName, newName, application)\\
		& changeReference(referncedClass, newReferencedClass) \\
	\textbf{Application Disassembly} 
		& removeProperty(className, propertyName,\\ &  \hspace{0.5in} application) \\
		& removeAssociation(className, associationName, \\ &  \hspace{0.5in}application) \\
		& removeClass(className, application)\\
	\textbf{Copy Structure} 	 
		& copyProperty(sourceClass, property, targetClass,\\ &  \hspace{0.5in} newName, application)  \\
		& copyClass(class, newName, application) \\
	\hline
	\end{tabular}

\end{table}

@@

\begin{align}
&	\mathbf{Database} = ( Table*, Row* )\\
&	\mathbf{Table} = (label, primaryKey, Column*, ForeignKey*)\\
&	\mathbf{Column} = (label, DbType, DefaultValue, Constraint*) \\
&	\mathbf{ForeignKey} = (Label, Label, Constraint*) \\
&	\mathbf{PrimaryKey} =  ( Label, Sequence ) \\
&	\mathbf{DbType} = DBSTRING \; | \; DBINT\\
&	\mathbf{Constraint} = NOTNULL \; | \; UNIQUE 
\end{align}

@@

A database consists not only from schema but also from data which are represented as rows in a table. A table row in our model is a tuple consisting of value pairs, which represents concrete value of concrete column or key. Each row contains a name of a table it belongs to.
\begin{align}
&	\mathbf{Row} = (Label, N, Pair*) \\
&	\mathbf{Pair} = (Label, Value) 
\end{align}

@@

\subsubsection{Database Manipulation}
\label{sec:db-evolution}
The database consists of two parts - from database schema which defines the structure and from stored data - hence the transformations have to consider both parts. The transformation for manipualtion of structure has similar conditions to the evolution of application application. The transformations for data manipulation are inspired by the SQL language. The basic transformations are in Table \ref{tab:db-basic-evolution}. 
\begin{table}
\caption{The transformations for database evolution.}
	\label{tab:db-basic-evolution}
\centering
\setlength\tabcolsep{0.5em}
	\begin{tabular}{ll}
	\hline
	Type of transformation & Defined transformations \\
	\hline
	\\[-2ex] Database Creation
	& addColumn(tableName, column, database) \\
	& addForeignKey(tableName, key, database) \\
	& addTable(table, database)\\
	Database Alternation
	& alterColumnName(tableName, oldName, \\ & \hspace{0.5in}newName, dataabse) \\
	& alterForeignKeyName(tableName, oldName, \\ & \hspace{0.5in}newName, dataabse) \\
	& alterTableName(oldName, newName, dataabse) \\
	Database Disasembling
	& dropColumn(tableName, columnName, database) \\
	& dropForeignKey(tableName, keyName, database) \\
	& dropTable(tableName, database) \\
	Data Manipulation
	& select(tableName, rowId, database) \\
	& select(tableName, database) \\
	& insert(row, database) \\
	& insert(rowId, value, database) \\
	& delete(tableName, rowId, database) \\
	& delete(tableName, rowId, columnName, database) \\
	Copy Structure
	& copyColumnStructure(sourceTableName,\\ &  \hspace{0.5in} targetTableName,  sourceColumnName, \\ &  \hspace{0.5in} newName, database) \\
	& copyTableStructure(sourceTable, newTableName, \\ &  \hspace{0.5in} database) \\
	Copy Strucutre and Values
	& copyColumn(sourceTable, targetTable, \\
	& \hspace{0.5in} sourceColumnName, targetColumnName, \\ 
	& \hspace{0.5in} mapping, database) \\
	& copyTable(sourceTable, newName, database) \\
	\hline
	\end{tabular}
\end{table}
The set of transformations for database manipulation is limited in contrast with the SQL language it is because mentioned transformation are necessary for application of data evolution on database level.

@@

The mapping between instances of the classes (stored data) to provides information which value from the source table rows will be copied to which row in target table. The value of a column is then inserted for each row into the copied column during $copyColumn$ transformation. 
The mapping is usually defined as: 
\begin{align}
&	mapping =  (r_1, r_2)*,\\
& 	r_1 \in select(ORM(sourceName), database) \cup \varnothing, \\
& 	r_2 \in select(ORM(targetName), database) \cup \varnothing
\end{align}
The mapping is not limited to the bijection and it is possible to not include a row into a mapping - this has to be verified because of possible $NOTNULL$ constraints on a target column, as well as the $UNIQUE$ constraint has to be verified because the mapping can violate it in general. On the other hand the benevolent mapping help with creating various kinds on advanced evolutionary transformations. The mapping can cause data losses in some cases if used as a part of advanced refactoring. This situation is discussed later on.

\paragraph{Data Secure Disasembling of a Software} The transformation for disassembly the database can have fatal impact on the data preservation, therefore the set of transformations is extended by a data secure transformations for database disassembling. These transformations are not part of the SQL standard, although they can be implemented as database functions. These transformations creates a secure way to disassembly the database as they drops empty structural elements only. These transformations are in table \ref{tab:db-adv-evolution}. 
\begin{table}
\centering
\caption{The data secure transformations for database evolution which can be created as SQL functions.}
	\label{tab:db-adv-evolution}
	\begin{tabular}{ll}
	\hline
	Type of transformation & Defined transformations \\
	\hline
	\\[-2ex] 
	Data Secure Database Disassembling &
	dropEmptyColumn(tableName, columnName, \\ & \hspace{0.25in} database) \\
	& dropEmptyForeignKey(tableName, keyName, \\ & \hspace{0.25in} database) \\
	& dropEmptyTable(tableName, database) \\
	\hline
	\end{tabular}
\end{table}

@@

\paragraph{Basic Software Evolution} The basic evolutionary transformations of the software are based on basic evolution of an application. These transformation cannot be  mapped directly on the evolutionary transformations of the database, because properties and associations can be mapped as a column (foreign key respectively) or as a table, this has to be reflected during the evolution as shown on the example of creating a new property: 
\begin{align}
& \Phi(newProperty(className, property, application)) = \nonumber \\
& \;\;\; =  \begin{cases}
  addColumn(ORM(className), ORM(property), database) \\\mathbf{if} \; cardinality(property \leq 1)  \\\\
  addTable(ORM(property), database) \\
  \mathbf{if} \; cardinality(property > 1)  
 \end{cases}
\end{align}

@@

\begin{table}
	\caption{The mapping between evolution of the application to the database transformations.}
	\label{tab:sw-basic-evolution}
\centering
	\begin{tabular}{ll}
		\hline
	Application Transformation & Database Transformation \\
	\hline
	\\[-2ex] 
	newProperty & \textbf{if} $cardinality \leq 1$ \textbf{then} addColumn \textbf{else} addTable \\

	newAssociation & \textbf{if} $cardinality \leq 1$ \textbf{then} addColumn  \textbf{else} addTable \\

	newClass & addTable \\
	renameProperty & \textbf{if} $cardinality \leq 1$ \textbf{then}  renameColumn  \textbf{else} renameTable \\

	renameAssociation & \textbf{if} $cardinality \leq 1$ \textbf{then} renameForeignKey \\ & \hspace{1in} \textbf{else} renameTable \\

	renameClass & renameTable \\
	removeProperty & \textbf{if} $cardinality \leq 1$ \textbf{then} removeColumn  \\ & \hspace{1in}\textbf{else} removeTable \\

	removeAssociation & \textbf{if} $cardinality \leq 1$ \textbf{then}  removeForeignKey \\ & \hspace{1in} \textbf{else} removeTable \\

	removeClass & removeTable\\
	\hline
	\end{tabular}
\end{table}

@@

The  $copyProperty$ is the first transformation when an additional information has to be added to the usual code refactoring. It is because the $copyColumn$ transformation needs one more information to succeed - there has to be provided mapping between the source and target table to assure the data information consistency. %Moreover the source and target should have the same structure, type and constraints.
\begin{align}
& \mathbf{copyProperty}(sourceClass, propertyName, targetClass, \nonumber  \\
& \; \; \; mapping, software) 
\end{align}
The transformation is interpreted for both software components - in case of the application a new property is added:
\begin{align}
& \Psi(copyProperty(sourceClass, property, targetClass, mapping, \nonumber  \\ 
&  \; \; \; software)) = newProperty(targetClass, property, application) 
\end{align}
in case of database the copy of a column or table is created according to property's cardinality and then values are copied:
\begin{align}
& \Phi(copyProperty(sourceClass, property, targetClass, newName, \nonumber \\ 
&  \; \; \; mapping, software)) = \begin{cases}
 coypColumn(ORM(sourceClass),  \\ \;\;\; ORM(targetClass),  \\ \;\;\;alterColumnName(ORM(property), \\ \;\;\; ORM(newName)), mapping, database) \\ \; \mathbf{if} \; cardinality(property) \leq 1 \\\\
 copyTable(ORM(Property), \\ \;\;\; ORM(newName))
 \end{cases}
\end{align}

@@

\paragraph{Move Property}
The $moveProperty$ transformation is based on the $copy$ $Property$ transformation followed by the $removeProperty$ so its construction is easy. On the other hand special attention has to be paid to the provided mapping of instances, because it can cause data loses. The ideal case is when the mapping is a bijection among instances, then the transformation can not cause the lost of data. On the other hand the situation when the mapping does not provides pairs for all rows in a source table can lead to data loss. 
\begin{align}
& moveProperty(sourceClass, property, targetClass, newName,\nonumber  \\
& \;\;\; mapping, software) = removeProperty(sourceClass, property, \nonumber \\
& \;\;\; CopyProperty(sourceClass, property, targetClass, \nonumber \\ 
& \;\;\;newName, mapping, software))
\end{align}

\paragraph{Merge Classes}
There occur situations when there is a need to merge two classes and the inlining transformation cannot be used because of name collisions among properties in such a case the $merge$ transformation can be used. It can be used in case structure of both classes (respectively tables) are equal. The transformation creates new rows in the new table - the primary keys from origin instances are replaced by a new ones.
\begin{align}
& merge(firstClass, secondClass, newClassName, software) = \nonumber \\
& \; \; \; = removeClass(secondClass, removeClass(firstClass,\nonumber  \\
& \; \; \; changeReference(secondClass, newClassName, \nonumber \\
& \; \; \; changeReference(firstClass, newClassName,\nonumber  \\
& \; \; \; copyClass(firstClass, newClassName, software))))))
\end{align}

\paragraph{Extract and Inline Class}
The extract and inline class are two opposite transformations. First of them extract a class from an existing class, second of them moves data from one class into another.
\begin{align}
& extract(sourceClass, newClass, property*, software) = \nonumber  \\ 
& \; \; \; = moveProperty(sourceClass, newClass, property*, \nonumber \\ 
& \; \; \; addClass(newClass, software))
\\ \nonumber  \\
& inline(targetClass, classToInline, software)) =  \nonumber \\
& \; \; \; = removeClass(classToInline, moveProperty(classToInline, \nonumber \\
& \; \; \; targetClass, properties(classToInline), software))
\end{align}

@@

\paragraph{ORM Frameworks} There are many object-relational mapping frameworks available for developers, and some of them provide tools for database migration. Hibernate \cite{Hibernate} is one of the most popular ORM frameworks in the Java community which is capable of creating a new table or adding a new column according to a change in the application. Active Record \cite{Active_Record} is an ORM framework in the Ruby on Rails environment. Since its first version it has contained support for database evolution according to the create-update-delete principle, in the form of so-called migrations \cite{Rails:Migrations} which can be extended by adding user SQL commands. Entity Framework \cite{Entity_Framework} is Microsoft's ORM solution for the .NET platform. Its capabilities of data evolution support are similar to those of Active Record. Neither of the frameworks is capable of automatization of complex refactoring cases.

\paragraph{Tools for Database Evolution} The MeDEA project \cite{Meda:Main} offers a tool for evolution of both database schema and stored data based on model-driven approach. The project DB-MAIN \cite{Hick:DBMAIN} provides a MDD approach to data evolution on all the levels of software we do. The project DB-MAIN is well documented formally. The PRISM is a research project for data management under schema evolution \cite{PRISM} in contrast with our proposal, and with MeDEA, it extends the SQL command set by so-called schema modification operators which implements the schema evolution.  Very promising solution is Liquibase \cite{Liquibase} a tool for database refactoring and evolution. It is capable to migrate both database schema and stored data. The evolution is described in a form of a XML document which can be interpreted on various databases.

The difference between these frameworks and our proposal is that each has a different focus. All mentioned projects are aimed to  be used by database administrators; whereas our focus is on entities evolution, which is then propagated to a database with an emphasis on automatization. Our goal is to hide the entire database level from our users.

@@

\bibliographystyle{splncs03}      % basic style, author-year citations
\bibliography{references}   % name your BibTeX data base
